Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
modified specs

In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$. When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}. The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not. The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:

\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
\end{align*}

The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (refer to line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  \texttt{t}. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{t}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
				\\ 
				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{t}
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{findNext(pn, k)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{res n'.}
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{t} \mapsto \texttt{(k', v', l, r)} \  \ast \\ 
				\ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the lock-coupling and give-up templates}
	\label{fig:findNext_lock}
\end{figure}

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{t' l r lk1 lk2}.
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
	\label{fig:insertOp_lock}
\end{figure}

@@

\begin{figure}
	\centering
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n} \ \ast \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}
		\end{array}
		\right\}
	}
	\\ 
	\texttt{inRange(pn, k)} 
	\\
	{\color{blue}
		\left\{\exists \  \texttt{res.}
		\begin{array}{c}
			 \ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}\  \\ 
			 \ \ast  \ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
		\end{array}
		\right\}
	}
\end{mathpar}
\caption{Specification of \texttt{inRange} for the give-up template}
\label{fig:inRange_giveup}
\end{figure}

\begin{figure}
	\centering
	\begin{mathpar}
	%spec of insertOp
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n} \ \ast \texttt{n->t} \  \ast \texttt{n->min} \ \ast \texttt{n->max}
		\end{array}
		\right\}
	}
	\\
	\texttt{insertOp(pn, k, v)}
	\\
	{\color{blue}
	\left\{\exists \ \texttt{t' l r lk1 lk2}. 
	\begin{array}{c}
		\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
		\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}
	\end{array}
	\right\}
	}
\end{mathpar}
\caption{Specification of \texttt{insertOp} for the give-up template}
\label{fig:insertOp_giveup}
\end{figure}

>>>>

In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$. When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}. The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not. The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
\end{align*}
The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (refer to line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  \texttt{m}. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{m}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
				\\ 
				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn}\mapsto\texttt{n} \ast \texttt{n}\mapsto\texttt{t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{findNext(pn, k)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{res n'.}
			\begin{array}{c}
				\texttt{pn}\mapsto\texttt{n} \ast \texttt{n}\mapsto\texttt{t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \\\  
				\ast \ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the lock-coupling and give-up templates}
	\label{fig:findNext_lock}
\end{figure}

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{t' l r lk1 lk2}.
			\begin{array}{c}
				\ \texttt{pn} \mapsto \texttt{n} \ \ast \ \texttt{n}\mapsto \texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
	\label{fig:insertOp_lock}
\end{figure}

@@

\begin{figure}
	\centering
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn}\mapsto\texttt{n} \ \ast \texttt{n}\mapsto\texttt{t} \ \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n->max}
		\end{array}
		\right\}
	}
	\\ 
	\texttt{inRange(pn, k)} 
	\\
	{\color{blue}
		\left\{\exists \  \texttt{res.}
		\begin{array}{c}
			 \ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \ \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}\  \\ 
			 \ \ast  \ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
		\end{array}
		\right\}
	}
\end{mathpar}
\caption{Specification of \texttt{inRange} for the give-up template}
\label{fig:inRange_giveup}
\end{figure}

\begin{figure}
	\centering
	\begin{mathpar}
	%spec of insertOp
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t} \  \ast \ \texttt{n}\mapsto\texttt{min} \ \ast \ \texttt{n}\mapsto\texttt{max}
		\end{array}
		\right\}
	}
	\\
	\texttt{insertOp(pn, k, v)}
	\\
	{\color{blue}
	\left\{\exists \ \texttt{t' l r lk1 lk2}. 
	\begin{array}{c}
		\ \texttt{pn}\mapsto\texttt{n} \ \ast \ \texttt{n}\mapsto\texttt{t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \\ 
		\ \ast \ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}
	\end{array}
	\right\}
	}
\end{mathpar}
\caption{Specification of \texttt{insertOp} for the give-up template}
\label{fig:insertOp_giveup}
\end{figure}
